---
title: "Bayesian statistics with R (gimenez)"
---

<https://oliviergimenez.github.io/bayesian-stats-with-R>

<https://github.com/oliviergimenez/bayesian-stats-with-R>

```{r}
library(tidyverse)
library(magrittr)
library(bayesrules)
theme_set(theme_bw())
set.seed(2020)
```

## Binomial MLE

```{r}
#| warning: false

data = c("x" = 3, "size" = 10)

negloglik <- function(p, data) {
  -dbinom(x = data["x"], size = data["size"], prob = p, log = TRUE)
}

negloglik(p = 0.3, data = data)

fit = optimize(f = negloglik, interval = c(0, 1), data = data)

fit
```

## Normal MLE

```{r}
n = 100

height = rnorm(n = n, mean = 170, sd = 10)

height
```

```{r}
#| warning: false
negloglik <- function(theta, data) {
  mu = theta[1]
  sigma = theta[2]
  x = data
  -sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}

negloglik(theta = c(150,1), data = height)

fit = optim(fn = negloglik, par = c(1,1), data = height)

fit
```

```{r}
summary(lm(height ~ 1, data = tibble(height)))
```

## Bayesian inference

57 sick mice were injected with medicine and 59 were injected with vehicle.

Of the 57 mice injected with medicine, 38 died and 19 survived.

Of the 59 mice injected with vehicle, 38 dies and 21 survived.

Does the medicine have an effect on survival?

| Treatment | Died | Survived | Total | Survival ($\pi$) |
|-----------|------|----------|-------|------------------|
| Medicine  | 38   | 19       | 57    | 33%              |
| Vehicle   | 38   | 21       | 59    | 36%              |

Our generative model is a binomial experiment (assuming independent and identically distributed observations, i.e., random draws from a binomial distribution).

MLE of survival is $\hat{\theta} = \frac{k}{n}$.

Bayesian inference begins with a prior for the survival parameter $\theta$.

We can use the beta distribution (bound from 0 to 1) as a prior for $\theta$.

```{r}
plot_beta(1, 1)
```

```{r}
plot_beta(1,2)
```

```{r}
plot_beta(2,1)
```

```{r}
plot_beta(2,2)
```

```{r}
plot_beta(10,10)
```

```{r}
plot_beta(0.5,0.5)
```

### Prior probability (Beta)

$$
\theta \sim \mathcal{B}(a,b)
$$

$$
Pr(\theta) = \theta^{a-1}(1-\theta)^{b-1} 
$$

### Likelihood (Binomial)

$$
Pr(k \mid \theta) = {n \choose k} \theta^k (1 - \theta)^{n - k}
$$

### Posterior probability

$$
Pr(\theta \mid k) \propto {n \choose k} \theta^k (1 - \theta)^{n - k} \theta^{a-1}(1-\theta)^{b-1} 
$$

$$
Pr(\theta \mid k) \propto \theta^{(a+k)-1} (1 - \theta)^{(b+n-k)-1} 
$$

$$
\theta \mid k \sim \mathcal{B}(a+k, b+n-k)
$$

A beta prior with a binomial likelihood produces a beta posterior!

```{r}
plot_binomial_likelihood(y = 3, n = 10, mle = TRUE)
```

```{r}
plot_beta_binomial(1, 1, 3, 10)
```

```{r}
plot_beta_binomial(a = 2, b = 2, y = 3, n = 10)
```
