---
title: "Introduction to modern statistics"
format:
  html:
    df-print: paged
---

```{r}
#| output: false
library(tidyverse)
library(magrittr)
library(janitor)
library(openintro)
```

## Introduction to data

Statistics is the study of how best to collect, analyze, and draw conclusions/learn from data.

We start by writing the main question researchers hope to answer:

> Does the use of stents reduce the risk of stroke?

Researchers randomly assigned 224 patients to the treatment group and 227 to the control group. In this study, the control group provides a reference point against which we can measure the medical impact of stents in the treatment group.

Researchers studied the effect of stents at two time points: 30 days after enrollment and 365 days after enrollment.

```{r}
stent30
```

```{r}
stent365
```

```{r}
rbind(stent30 %>% mutate(time = "30"), stent365 %>% mutate(time = "365")) %>%
  tabyl(group, outcome, time) %>%
  adorn_totals("row") %>%
  adorn_percentages("all") %>%
  adorn_pct_formatting(digits = 0) %>%
  adorn_ns %>%
  adorn_title
```
